---
title: "Initial exploration of Great Lakes historical water levels"
output: html_notebook
---

```{r libs, warning=FALSE, message=FALSE}
library(readr)
library(lubridate)
library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)

```
Dashboard page: https://www.glerl.noaa.gov/data/dashboard/GLD_HTML5.html

Data download page: https://www.glerl.noaa.gov/data/dashboard/data/

Get all the data as a zip file:
https://www.glerl.noaa.gov/data/dashboard/GLDData.zip

Great Lakes Environmental Research Lab
https://www.glerl.noaa.gov/

Great Lakes Water Level Overview and Observations tab
https://www.glerl.noaa.gov/data/wlevels/levels.html#overview
https://www.glerl.noaa.gov/data/wlevels/levels.html#observations




We'll read in the data with readr. The raw data file was obtained from:

http://lre-wm.usace.army.mil/ForecastData/GLHYD_data_metric.csv

Unfortunately, it appears the above file is only updated annually.
We will need to reconstruct it from the raw data files if we want
to analyze recent 2019 data.

```{r, readGLHYD, message=FALSE}
# Top 12 lines are meta-data
GLHYD_data_metric <- read_csv("GLHYD_data_metric.csv", skip = 12)
```

We can download csv files from NOAA at the following location:

```{r}
data_loc <- "https://www.glerl.noaa.gov/data/dashboard/data/levels/1918_PRES/"
```

Read in monthly water level data for each lake. Top two lines are metadata. Here's what a little bit of the Lake Superior file, `superior1918.csv`, looks like.

    Lake Superior:, Monthly Lake-Wide Average Water Levels (1918 - Present)
    Source:, NOAA/NOS; CHS
    year,jan,feb,mar,apr,may,jun,jul,aug,sep,oct,nov,dec
    1918,183.25,183.2,183.17,183.14,183.22,183.34,183.4,183.43,183.45,183.44,183.46,183.44
    1919,183.38,183.32,183.28,183.31,183.38,183.45,183.48,183.47,183.46,183.42,183.43,183.39
    1920,183.29,183.27,183.28,183.36,183.43,183.52,183.58,183.6,183.54,183.5,183.43,183.36



```{r, readcsv, message=FALSE}
miHuron1918 <- read_csv(str_c(data_loc, "miHuron1918.csv"), skip = 2)
ontario1918 <- read_csv(str_c(data_loc, "ontario1918.csv"), skip = 2)
erie1918 <- read_csv(str_c(data_loc, "erie1918.csv"), skip = 2)
superior1918 <- read_csv(str_c(data_loc, "superior1918.csv"), skip = 2)
clair1918 <- read_csv(str_c(data_loc, "ontario1918.csv"), skip = 2)
```

Looking at the structure of one of these tables reveals that it is in
wide format with the months as columns.

```{r}
head(miHuron1918)
str(miHuron1918)
```

Notice that the data is in
*wide format* with the months as columns. In order to facilitate plotting with ggplot2, we are going to need to reshape this data into *long format*. Each row will be a single monthly reading and their will be a month column. This could be done with the `melt()` function from the `reshape2` package or `gather()` from the `tidyr` package. Let's use the `tidyr::gather` function. 

```{r}
miHuron1918_long <- gather(miHuron1918, "month", "mihuron", 2:13)
ontario1918_long <- gather(ontario1918, "month", "ontario", 2:13)
erie1918_long <- gather(erie1918, "month", "erie", 2:13)
superior1918_long <- gather(superior1918, "month", "superior", 2:13)
clair1918_long <- gather(clair1918, "month", "clair", 2:13)
```

Take a peek at the Lake Huron data in long format.

```{r}
head(miHuron1918_long)
```

The months are three character strings. Let's create a date column that we can
use for joining the individual tables as well as proper sorting.

```{r}
miHuron1918_long$date <- as.POSIXct(str_c(miHuron1918_long$month, " 1, ",
                                                miHuron1918_long$year),
                                          format="%b %d, %Y")

ontario1918_long$date <- as.POSIXct(str_c(ontario1918_long$month, " 1, ",
                                                ontario1918_long$year),
                                          format="%b %d, %Y")

erie1918_long$date <- as.POSIXct(str_c(erie1918_long$month, " 1, ",
                                                erie1918_long$year),
                                          format="%b %d, %Y")

superior1918_long$date <- as.POSIXct(str_c(superior1918_long$month, " 1, ",
                                                superior1918_long$year),
                                          format="%b %d, %Y")

clair1918_long$date <- as.POSIXct(str_c(clair1918_long$month, " 1, ",
                                                clair1918_long$year),
                                          format="%b %d, %Y")
```

```{r}
head(miHuron1918_long)
```

Drop the year and month columns as they are no longer needed and move the
date to the left most column.

```{r}
miHuron1918_long <- miHuron1918_long[, c(4,3)]
ontario1918_long <- ontario1918_long[, c(4,3)]
erie1918_long <- erie1918_long[, c(4,3)]
superior1918_long <- superior1918_long[, c(4,3)]
clair1918_long <- clair1918_long[, c(4,3)]
```


Now we can join the long format tables together using the new date field.

```{r}
lakes_long <- merge(miHuron1918_long, ontario1918_long, by = "date")
lakes_long <- merge(lakes_long, erie1918_long, by = "date")
lakes_long <- merge(lakes_long, superior1918_long, by = "date")
lakes_long <- merge(lakes_long, clair1918_long, by = "date")
```

```{r}
head(lakes_long)
```

Get dataframe into "tidy" format by creating a `lake` field and gathering (melt) the lake fields.

```{r}
lake_level <-  gather(lakes_long, 2:6, key = "lake", value = "waterlevelm")
lake_level <- lake_level %>% 
  arrange(lake, date)
head(lake_level)
```

Clear up the workspace by getting rid of unneeded dataframes.

```{r}
rm(miHuron1918, ontario1918, erie1918, superior1918, clair1918)
rm(miHuron1918_long, ontario1918_long, erie1918_long, superior1918_long, clair1918_long)
rm(lakes_long)
```

If we plot all the lake level time series on one plot, we can see the differences in levels between lakes but the intra-lake variation is hidden by the scale.

```{r, ts1, warning=FALSE}
ggplot(lake_level) + geom_line(aes(x=date, y = waterlevelm, colour = lake)) +
  ylab("Water level (m)")
```

Where did Lake St. Clair go? Probably hidden under Lake Ontario.

```{r, ts1b, warning=FALSE}
lake_level %>% 
  filter(lake == 'clair' | lake == 'ontario') %>% 
ggplot() + geom_line(aes(x=date, y = waterlevelm, colour = lake)) +
  ylab("Water level (m)")
```

Hmm, are the measurements identical for Lake St. Clair and Lake Ontario?

```{r}
all(lake_level[lake_level$lake == 'ontario' & !is.na(lake_level$waterlevelm), 3] == lake_level[lake_level$lake == 'clair' & !is.na(lake_level$waterlevelm), 3])
```

Ok, let's drop Lake St. Clair.

```{r}
lake_level <- lake_level %>% 
  filter(lake != 'clair')
```

Let's find the latest month for which we have data and save a version of this dataframe and tag it by the month.

Here's how we can do this with base R.

```{r lastmnth_base}
last_month <- max(lake_level[!is.na(lake_level$waterlevelm),"date"])
```

And, here's a `dplyr` approach. Note the use of `pull()` to convert the the resulting 1x1 dataframe into a value.

```{r lastmnth_dplyer}
last_month_d <- lake_level %>% 
  filter(!is.na(waterlevelm)) %>% 
  select(date) %>% 
  summarize(whichmonth = max(date)) %>% 
  pull()
```

Save as an rds file.

```{r}
rdsname <- paste0("data/lake_level_", year(last_month), strftime(last_month,"%m"), ".rds")

saveRDS(lake_level, rdsname)
```


```{r fig.width=16, fig.height=15, warning=FALSE}
ts_all <- ggplot(lake_level) + geom_line(aes(x=date, y = waterlevelm, colour = lake)) + ylab("Water level (m)") +
  facet_grid(lake~., scales = "free")

ts_all
```

Compute water levels summary stats using `dplyr`.

```{r}
lake_level_stats <- lake_level %>%
  group_by(lake) %>%
  summarize(
    mean_level = mean(waterlevelm, na.rm = TRUE),
    sd_level = sd(waterlevelm, na.rm = TRUE),
    cv_level = sd_level / mean_level,
    min_level = min(waterlevelm, na.rm = TRUE),
    max_level = max(waterlevelm, na.rm = TRUE),
    p05_level = quantile(waterlevelm, 0.05, na.rm = TRUE),
    p95_level = quantile(waterlevelm, 0.95, na.rm = TRUE)
  )

lake_level_stats
```


https://drsimonj.svbtle.com/plotting-individual-observations-and-group-means-with-ggplot2

Let's try to combine time series with lines from the stats dataframe just created. Start with
just one lake and then we can try with faceted plot by lake.



```{r}
lake_level_huron <- lake_level %>%
  filter(lake == 'mihuron')

lake_level_stats_huron <- lake_level_stats %>%
  filter(lake == 'mihuron')
```

```{r tshuron, warning=FALSE}
ggplot(lake_level_huron, aes(x = date, y = waterlevelm)) +
  geom_line() + geom_hline(yintercept = lake_level_stats_huron$mean_level)
```


Tried and true approach is to add columns to original dataframe with repeated values or create
dataframe by date with repeated values (and then try to overlay)


```{r}
lake_level_ts_stats <- merge(x = lake_level, y = lake_level_stats) %>%
  arrange(lake, date)
```


Add overall mean and percentile bands.

```{r ig.width=16, fig.height=15, warning=FALSE}
ggplot(lake_level_ts_stats, aes(x = date, y = waterlevelm, colour = lake)) +
  facet_grid(lake ~ ., scales = "free") +
  geom_line() + 
  geom_line(aes(y = mean_level)) + 
  geom_line(linetype = "dashed", aes(y = p05_level)) + 
  geom_line(linetype = "dashed", aes(y = p95_level))
```

## Time series analysis in R

Quite a few packages. The [CRAN Task View: Time Series Analysis](https://cran.r-project.org/web/views/TimeSeries.html) page maintained by Rob J Hyndman is the place to start to see what's available.

zoo and xts seem foundational.

https://cran.r-project.org/web/packages/hydroTSM/index.html

Let's add moving average using `tidyquant` library (which wraps a bunch of quant finance
relevant libraries).


```{r lib_tidyquant, warning=FALSE, message=FALSE}
library(tidyquant)
```

```{r ts_SMA, fig.width=16, warning=FALSE}
lake_level_ts_stats %>% 
ggplot(aes(x = date, y = waterlevelm, colour = lake)) +
  facet_grid(lake ~ ., scales = "free") +
  geom_line() + 
  geom_line(aes(y = mean_level)) + 
  geom_line(linetype = "dashed", aes(y = p05_level)) + 
  geom_line(linetype = "dashed", aes(y = p95_level)) + 
  geom_ma(ma_fun = SMA, n=12, colour = "blue", linetype = "solid", size = 1)
```



